---
title: "Publishing a result from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publishing a result from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A Neuropublish publication is a *bundle*: a directory with a `manifest.json`
written to a public schema and the NIfTI files it names. The manifest says
what each volume *is* (which analysis, which estimand, which measure, on which
grid), what the viewer should show first, and where the numbers came from.
Transport is the job of the reference command-line tool `npub`; this package
builds the manifest as plain R lists and wraps the tool. Neither the bundle nor
`npub` needs R.

This vignette builds the same bundle the Julia conformance producer writes
(`modules/conformance/julia/producer.jl`) from `neuroim2` volumes, validates
it, and publishes it when a server is available.

```{r setup}
library(neuropublish)
```

```{r cli-gate, include = FALSE}
# The chunks that run the CLI are opt-in: NPUB_TESTS=1 plus a resolvable `npub`.
np_cli_available <- function() nzchar(Sys.getenv("NPUB_TESTS")) && np_has_npub()
```

## The volumes

`np_synthetic_volumes()` evaluates the producer's formulas on its
16 × 16 × 12 grid (2 mm, origin −15, −15, −11), rounded to float32, so every
voxel value is known: a T1-like underlay and an effect, *t*, and *z* map.

```{r volumes}
vols <- np_synthetic_volumes()
names(vols)
round(vols$`speech-t`[8, 8, 6], 4)
neuroim2::space(vols$`speech-t`)
```

## The manifest

A manifest starts with a title, a synopsis and its sensitivity (only
`group-level` revisions can be link-shared). Each builder returns a plain list
that serializes to exactly one record of the schema.

```{r manifest}
m <- np_manifest(
  "Julia producer — synthetic speech group model",
  "The Julia conformance bundle, rebuilt from R.",
  sensitivity = "group-level"
)
```

### Domain

The grid is described once, as a `volume-grid` domain. The open *descriptor*
carries shape and affine (from `neuroim2::trans()`); the exact *key* carries
the ADR 0005 structural fingerprint, a SHA-256 over a documented binary
preimage, which lets the server tell whether two results share a grid without
parsing anyone's payload. R, Julia and Scala compute the same bytes:

```{r domain}
dom <- np_domain_volume(neuroim2::space(vols$t1), id = "grid-2mm")
dom$key$structuralFingerprint
identical(
  dom$key$structuralFingerprint,
  "sha256:aa1d8cdb290b7739ff086b1f380552f59366c8671dc33425716abd5f2bce7687"
) # the value in fixtures/julia/manifest.json
m <- np_add(m, "domains", dom)
```

### Assets

`np_asset_volume()` writes a `NeuroVol` to NIfTI in a staging directory (or
references an existing file) and records its `path`. Content identity
(`digest`, `size`) is filled in later by `npub pack`, which hashes the bytes.

```{r assets}
staging <- file.path(tempdir(), "vignette-staging")
m <- np_add(m, "assets", np_asset_volume("t1", vols$t1, staging = staging, catalog = "synthetic:r/t1"))
for (id in c("speech-effect", "speech-t", "speech-z")) {
  m <- np_add(m, "assets", np_asset_volume(id, vols[[id]], staging = staging))
}
m$assets[[1]]
```

### Analysis, fields, underlay

An analysis declares its estimands; a result field is one measure of one
estimand on one domain. The method is an *open record*: a schema reference
from your own namespace plus a payload the server keeps but does not
interpret. Measures come from `np_measure`. A `publishedDisplay` is a
recommendation for the viewer, never an inferential decision.

```{r analysis}
m <- np_add(m, "analyses", np_analysis("group-model", "Group model · speech (synthetic)",
  method = np_record("org.example.r/reducer/mean", "0.1", list(weights = "equal", inputs = 12)),
  sample_size = 12,
  estimands = list(np_estimand("speech", "speech coefficient", order = 1))
))
m <- np_add(m, "resultFields", np_field("speech-effect", "speech", np_measure$effect, "grid-2mm",
  representations = list(np_volume_rep("speech-effect")), order = 1
))
m <- np_add(m, "resultFields", np_field("speech-t", "speech", np_measure$t_statistic, "grid-2mm",
  representations = list(np_volume_rep("speech-t")), order = 2,
  published_display = np_display(np_threshold("two-sided", 2.5), np_window(-6, 6, centre = 0), "cold-hot")
))
m <- np_add(m, "resultFields", np_field("speech-z", "speech", np_measure$z_statistic, "grid-2mm",
  representations = list(np_volume_rep("speech-z")), order = 3,
  published_display = np_display(np_threshold("two-sided", 2.3), np_window(-5, 5, centre = 0), "cold-hot")
))
m <- np_add(m, "underlays", np_underlay("t1", "grid-2mm", "Synthetic T1"))
```

### Provenance and warnings

Provenance is a small graph of entities and activities; activities are open
records too (here the `fmrireg` analysis-receipt vocabulary the reference
bundle uses). A warning names what it concerns.

```{r provenance}
receipt <- function(subject, noise) {
  np_record("org.bbuchsbaum.fmrireg/analysis-receipt", "1.0",
    list(subject = subject, temporalNoise = noise, drift = "cosine-128s", hrf = "spmg1"),
    digest = "sha256:aa00000000000000000000000000000000000000000000000000000000000001"
  )
}
m$provenance <- np_provenance(
  entities = list(np_entity("raw", "Synthetic raw data")),
  activities = list(
    np_activity("first-level-01", receipt("01", "AR(2)")),
    np_activity("first-level-02", receipt("02", "AR(1)")),
    np_activity("denoise", np_record("org.example.r/denoise", "0.1",
      list(method = "wavelet", levels = 3, threshold = 0.05)
    ))
  ),
  edges = list(
    np_edge("raw", "first-level-01"), np_edge("raw", "first-level-02"),
    np_edge("first-level-01", "denoise"), np_edge("first-level-02", "denoise"),
    np_edge("denoise", "speech-t")
  )
)
m <- np_add(m, "warnings", np_warning("heterogeneous-temporal-noise",
  "First-level temporal noise models differ across inputs.",
  concerns = list(analysis = "group-model")
))
```

### Extension fields

The manifest is a list: add what your package needs. Unknown fields are kept
by the writer, by `npub pack`, and by the server (value preservation, ADR
0001).

```{r extension}
m$`x-r-client` <- list(version = "0.1", note = "unknown top-level extension")
```

## Writing, packing, validating

`np_write_bundle()` writes the *staging* bundle under the byte profile (UTF-8,
no byte-order mark, finite numbers only: a `NaN` is refused with its JSON
Pointer). `npub pack` then hashes the files into a normalized bundle, and
`npub validate` runs the server's admission on the result.

```{r write}
work <- file.path(tempdir(), "vignette-bundle")
unlink(work, recursive = TRUE)
staged <- np_write_bundle(m, file.path(work, "staging"))
list.files(staged, recursive = TRUE)
```

```{r inspect}
written <- np_read_manifest(staged)
str(written$domains[[1]]$key)
written$assets[[2]]
written$`x-r-client`
```

The remaining steps need `npub` (found through `NPUB_BIN`, the `PATH`, or a
Neuropublish checkout's `scripts/npub`). Running the tool is opt-in, so that
building this vignette — and `R CMD check` — never depends on a JVM or a
build: the chunks below run only when `npub` is available *and* `NPUB_TESTS=1`
is set. See "Running the integration tests" at the end.

```{r pack, eval = np_cli_available()}
packed <- np_pack(staged, file.path(work, "bundle.npub"))
packed$digest
packed$assets
problems <- np_validate(packed$dir)
nrow(problems) # zero rows: admitted
attr(problems, "digest")
```

A manifest that breaks a rule comes back as rows of JSON Pointer and message:

```{r invalid, eval = np_cli_available()}
bad <- np_read_manifest(packed$dir)
bad$resultFields[[1]]$estimand <- "no-such-estimand"
np_write_bundle(bad, file.path(work, "bad"))
np_validate(file.path(work, "bad"))
```

## Publishing

After a one-time `np_login(server)` (device flow: the CLI prints a URL and a
code you approve in any browser), `np_push()` uploads only the objects the
server lacks and commits one immutable revision. Batch jobs use a
project-scoped publisher credential in `NP_TOKEN` instead of a login. This
chunk runs only when `NPUB_SERVER` is set.

```{r push, eval = nzchar(Sys.getenv("NPUB_SERVER")) && np_cli_available()}
server <- Sys.getenv("NPUB_SERVER")
project <- Sys.getenv("NPUB_PROJECT", "rotman/sherlock")
r <- np_push(packed$dir, project, server, message = "R vignette")
r$revision
r$view_url
```

A second push with a parent that is no longer the head is rejected with the
current head; the error has class `np_stale_parent` and a `head` field:

```{r stale, eval = FALSE}
tryCatch(
  np_push(packed$dir, project, server, parent = "some-old-revision"),
  np_stale_parent = function(e) e$head
)
```

## One call: `np_publish()` and `as_neuropublish()`

`as_neuropublish()` is the generic a domain package implements: it returns a
`neuropublish_result` (a manifest whose assets carry `path`s, plus the asset
list). `np_publish()` is `as_neuropublish()` → write → pack → push. The
reference method for a named list of `NeuroVol`s builds the structure above
from a measure map:

```{r publish-list}
r <- as_neuropublish(vols,
  title = "Synthetic speech model", synopsis = "From a list of volumes.",
  measures = c("speech-effect" = np_measure$effect, "speech-t" = np_measure$t_statistic,
    "speech-z" = np_measure$z_statistic),
  estimand = "speech", estimand_label = "speech coefficient",
  underlay = "t1", underlay_label = "Synthetic T1", domain_id = "grid-2mm",
  staging = staging
)
r
```

```{r publish, eval = FALSE}
np_publish(vols, "rotman/sherlock", server = "http://127.0.0.1:8080",
  title = "Synthetic speech model", synopsis = "From a list of volumes.",
  measures = c("speech-t" = np_measure$t_statistic)
)
```

### What the domain packages' methods return

The methods live in the domain packages, not here. Each returns the same
`neuropublish_result`; what differs is what it knows:

* `as_neuropublish.fmrigds_result` (fmrigds): one analysis per reducer
  (`method = np_record("org.bbuchsbaum.fmrigds/reducer/<name>", ...)` with
  the reducer's options as payload, `sample_size` from the inputs), one
  estimand per contrast, and one field per emitted map with the right measure
  (`effect`, `standard_error`, `t_statistic`, `z_statistic`); a
  `np_warning()` for heterogeneous first-level models, and a `provenance`
  whose activities are the inputs' `fmrireg` receipts
  (`org.bbuchsbaum.fmrireg/analysis-receipt@1.0`) and whose edges lead from
  the raw dataset through the receipts and the reducer to the published
  fields.
* `as_neuropublish.rMVPA_result` (rMVPA): one analysis per model with the
  model specification as the method payload, estimands for the performance
  metrics, fields with `np_measure$accuracy` / `correlation` (or the package's
  own measure ids for what the protocol does not name), and provenance
  activities carrying the runtime context, session information, and Git
  identity rMVPA already records.
* `neuromosaic::publish()` calls `np_publish()` on the report's result and
  keeps the static Quarto output as a companion artifact, so a static export
  and the interactive publication come from one description.

Every method ends the same way:

```{r contract, eval = FALSE}
np_result(manifest) # manifest$assets entries carry `path`; np_publish() does the rest
```

## Running the integration tests

The package's own tests fall in two groups. The default group is pure R — the
builders, the byte profile, the fingerprint against the pinned reference and
Julia fixtures, and the parsing of the CLI's `--json` documents from fixture
strings. It needs nothing but the package's dependencies, and it is what
`R CMD check` runs:

```sh
R -q -e 'testthat::test_local("clients/r/neuropublish")'
```

The second group actually runs `npub` — packing, admission, the rotated-affine
key agreement — and is opt-in through `NPUB_TESTS=1`, because it needs the
built CLI (and, through `scripts/npub`, sbt and a JVM). From a Neuropublish
checkout:

```sh
sbt publisherCli/compile                       # once, so scripts/npub is fast
NPUB_TESTS=1 R -q -e 'testthat::test_local("clients/r/neuropublish")'
```

`npub` is located through `NPUB_BIN`, then the `PATH`, then `scripts/npub` of a
checkout above the working directory; set `NPUB_BIN` to test an installed
binary. The same `NPUB_TESTS=1` makes this vignette's `npub` chunks run.

The end-to-end test also publishes to a running control plane, so it needs a
server as well:

```sh
NP_DATA_DIR=$(mktemp -d) NP_PORT=8090 sbt backend/run &   # wait for /api/v1/health
NPUB_CONFIG_DIR=$(mktemp -d) scripts/npub login --server http://127.0.0.1:8090
NPUB_TESTS=1 NPUB_SERVER=http://127.0.0.1:8090 NPUB_PROJECT=rotman/sherlock \
  R -q -e 'testthat::test_local("clients/r/neuropublish", filter = "e2e")'
```
